Title: Metagenomic diversity analysis of metagenome-derived viral sequences from a Neotropical river

Description: This script performs taxonomic annotation, filtering, and alpha/beta diversity 
analysis of metagenomic sequencing data obtained from water column and 
sediment samples collected along a pollution gradient in the Virilla River, 
Costa Rica. Sampling was performed during the dry and rainy seasons.

Main Steps:
 1. Installation and loading of required R packages.
 2. Taxonomic assignment using local SQLite database and `taxonomizr`.
 3. Construction of `phyloseq` objects from OTU table, taxonomic assignments, 
    and metadata.
 4. Rarefaction of samples to normalize sequencing depth.
 5. Prevalence- and abundance-based filtering of taxa and samples.
 6. Subsetting of samples by habitat (water vs. sediment) and season (dry vs. rainy).
 7. Alpha diversity analysis using Observed richness, Shannon, and Inverse Simpson indices.
 8. Visualization of alpha diversity across sites and sample types.
 9. Statistical testing using Kruskal-Wallis and Dunn’s post-hoc tests.
 10.Ordination analysis on CLR-transformed microbial community data to visualize patterns in sample composition using Principal Component Analysis (PCA). 
 11. COGs categories visual analysis.
 
```{r install-packages, eval=FALSE}
BiocManager::install("phyloseq",force = TRUE)
install.packages("remotes", repos = "https://cloud.r-project.org")
remotes::install_github("vmikk/metagMisc", force = TRUE)
install.packages(
  "microViz",
  repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
)
BiocManager::install("metagenomeSeq")
install.packages("devtools", repos = "https://cloud.r-project.org")
devtools::install_github("braddmg/b2p/package")
```
 
 
```{r}
#Load required libraries
library("phyloseq")
library("phyloseq")
library("ggplot2")
library("vegan")
library("remotes")
library("metagMisc")
library("microbiome")
library("DESeq2")
library("taxonomizr")
library("tidyverse")
library(vegan)
library(picante)
library(ALDEx2)
library(metagenomeSeq)
library(HMP)
library(dendextend)
library(rms)
library(breakaway)
library(gridExtra)
library(ggplot2)
library(microViz)
library(patchwork)
library(tidyverse)
library(dplyr)
library(gridExtra)
library(cowplot)
library(forcats)
library(dplyr)
library(ggplot2)
library(forcats)
library(cowplot)

# Load the 'b2p' package, which provides tools for taxonomic assignment
library(b2p)
# Set up the local SQL taxonomic database by specifying its file path
setupdb(path = "C:/Users/DELL/Desktop/ManuscritoViroma")  # Replace with your actual path to 'accessionTaxa.sql'


#Download SQLite database for taxonomy annotation
#prepareDatabase('accessionTaxa.sql')

# Load taxonomic IDs from CSV
taxid <- read.csv("taxaid.csv")           
taxaid <- list(taxid$taxid)               # Extract 'taxid' column and convert to list
taxaid <- as.numeric(unlist(taxaid))      # Unlist and convert to numeric vector

# Get taxonomy information for each taxid
taxonomy <- getTaxonomy(
  taxaid,
  'accessionTaxa.sql',                    # SQLite database file for taxonomy lookup
  desiredTaxa = c("superkingdom", "phylum", "class", 
                  "order", "family", "genus", "species")  # Specify taxonomic levels
)

# Clean up row names (remove any whitespace)
row.names(taxonomy) <- gsub("[[:space:]]", "", row.names(taxonomy))

#Import OTU table and metadata
Data=read.table("viromes.csv", sep=",", row.names = 1, header=TRUE)
Data[is.na(Data)] <- 0

# Create OTU and taxonomy objects
OTU <- otu_table(Data, taxa_are_rows = TRUE)    
TAX <- tax_table(taxonomy)                    

# Create initial phyloseq object with OTU and taxonomy data
physeq <- phyloseq(OTU, TAX)                    
sample_names(physeq)                            
#taxa_names(physeq)                              

# Load sample metadata and format it to match the sample names in physeq
meta_data <- read.csv("mdata2.csv", sep = ",")  
sampledata <- sample_data(data.frame(
  meta_data,
  row.names = sample_names(physeq),
  stringsAsFactors = FALSE                      
))
#sampledata                                       

# Merge OTU, taxonomy, and metadata into a unified phyloseq object
physeq1 <- merge_phyloseq(physeq, sampledata)  
#sample_data(physeq1)                             

# Perform rarefaction to normalize sequencing depth across samples
p.rar2 <- rarefy_even_depth(
  physeq1,
  rngseed = 1,                                   # Set seed for reproducibility
  sample.size = min(sample_sums(physeq1)),       # Use minimum sequencing depth
  replace = FALSE                                 # Sample without replacement
)

# View metadata column names and rarefied phyloseq object summary
colnames(sample_data(p.rar2))  # Check available metadata variables
#p.rar2                         # View summary of the rarefied phyloseq object

# Filter taxa by prevalence across samples
# - prev.trh = 0.05: keep taxa present in at least 5% of the samples
# - threshold_condition = "OR": keeps taxa that meet *either* prevalence or abundance thresholds
# - abund.trh = NULL: skip abundance threshold for this step
# - abund.type = "total": considers total abundance per taxon
p.rar4 <- phyloseq_filter_prevalence(
  p.rar2,
  prev.trh = 0.05,
  abund.trh = NULL,
  threshold_condition = "OR",
  abund.type = "total"
)
#p.rar4  

# Further filter samples to retain only those with minimum total abundance of 10 reads
p.rar4.2 <- phyloseq_filter_sample_wise_abund_trim(
  p.rar4,
  minabund = 10
)
```


##### METADATA ######
-Sampling corresponds to the time of the year: 2-Dry season, 3-Rainy season.
-Type corresponds to the two types of samples: Water (free-living particles in the water column) and Sediment (particles settled to the bottom of the river)
-Site corresponds to three pollution gradients: Site 1=low contamination, Site 2=medium contamination, Site 3=high contamination.

```{r}
# Subset samples from the filtered phyloseq object (p.rar4.2) 

# Subset for the dry season samples (Muestreo == 2)
Tseca2 = subset_samples(p.rar4.2, Muestreo == 2)

# Subset for the rainy season samples (Muestreo == 3)
Tlluviosa2 = subset_samples(p.rar4.2, Muestreo == 3)

# Subset all water colum samples, then divide them by season
ColAgua2=subset_samples(p.rar4.2, Tipo == "water")
ColaAgua_Seca2 = subset_samples(ColAgua2, Muestreo == 2)        # Water samples from dry season
ColaAgua_Lluvioso2 = subset_samples(ColAgua2, Muestreo == 3)    # Water samples from rainy season

# Subset all sediment samples, then divide them by season
Sediment2=subset_samples(p.rar4.2, Tipo == "sediment")
Sediment2_Seca2 = subset_samples(Sediment2, Muestreo == 2)      # Sediment samples from dry season
Sediment2_Lluvioso2 = subset_samples(Sediment2, Muestreo == 3)  # Sediment samples from rainy season
```


Alfa diversity analysis 
```{r}

library(ggh4x) # Load ggh4x to set custom y-scales per panel

# Define shared y-axis limits for each metric across BOTH seasons
limits_observed   <- c(90, 230)
limits_shannon    <- c(3.9, 5.3)
limits_invsimpson <- c(0, 145)


# Define matching dark border colors
border_colors <- c("agua" = "#882255", "sedimento" = "#004D40")

# Globally set ggplot's default point size to small (e.g., size = 1 or 1.2)
update_geom_defaults("point", list(size = 1, alpha = 0.7))

# Create a richness plot (Observed, Shannon, InvSimpson) for dry season samples
DrySeason <- plot_richness(Tseca2, x = "Punto", measures = c("Observed", "Shannon", "InvSimpson")) +
  geom_boxplot(
    aes(fill = as.factor(Tipo), color = as.factor(Tipo)), # Added color mapping
    outlier.shape = NA, 
    position = position_dodge(width = 0.8),
    lwd = 0.4,       # Keeps borders crisp without overpowering the box
    alpha = 0.85
  ) +  
  
  scale_fill_manual(values = c("#CC79A7", "#009E73"), name = "Tipo") + 
  
  # Sets the custom border colors and hides duplicate legend entries
  scale_color_manual(values = border_colors, guide = "none") +
  
  scale_x_discrete(labels = c("Sitio 1" = "Site 1", "Sitio 2" = "Site 2", "Sitio 3" = "Site 3")) +
  facet_wrap(~variable, nrow = 1, scales = "free_y") + 
  
  # Standardize scales across columns for Dry Season
  facetted_pos_scales(
    y = list(
      variable == "Observed"   ~ scale_y_continuous(limits = limits_observed),
      variable == "Shannon"    ~ scale_y_continuous(limits = limits_shannon),
      variable == "InvSimpson" ~ scale_y_continuous(limits = limits_invsimpson)
    )
  ) +
  
  theme_bw(base_size = 14) +   
  theme(
    panel.grid = element_line(), 
    axis.line = element_line(color = "black"), 
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 14, angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    legend.position = "none",
    legend.title = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  ggtitle("Dry season") +
  labs(y = "Alpha diversity measure", x = "Site") 

# Display the plot
DrySeason


# 2. Add color mapping to geom_boxplot and scale_color_manual to the plot
RainySeason <- plot_richness(Tlluviosa2, x = "Punto", measures = c("Observed","Shannon", "InvSimpson")) +
  
  geom_boxplot(
    aes(fill = as.factor(Tipo), color = as.factor(Tipo)), # Added color mapping
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    lwd = 0.4,       
    alpha = 0.85
  ) +  
  
  scale_fill_manual(
    values = c("#CC79A7", "#009E73"), 
    name = "Tipo",
    labels = c("sedimento" = "sediment", "agua" = "water")
  ) + 
  
  # Sets the custom border colors and hides duplicate legend entries
  scale_color_manual(values = border_colors, guide = "none") + 

  scale_x_discrete(labels = c("Sitio 1" = "Site 1", "Sitio 2" = "Site 2", "Sitio 3" = "Site 3")) +
  facet_wrap(~variable, nrow = 1, scales = "free_y") + 
  
  facetted_pos_scales(
    y = list(
      variable == "Observed"   ~ scale_y_continuous(limits = limits_observed),
      variable == "Shannon"    ~ scale_y_continuous(limits = limits_shannon),
      variable == "InvSimpson" ~ scale_y_continuous(limits = limits_invsimpson)
    )
  ) +
  
  theme_bw(base_size = 14) +  
  theme(
    panel.grid = element_line(), 
    axis.line = element_line(color = "black"), 
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 14, angle = 90, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    legend.position = "bottom",
    legend.title = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  ggtitle("Rainy season") +
  labs(y = "Alpha diversity measure", x = "Site") 

RainySeason

plotTipoMuestreoPunto2 <- (DrySeason / RainySeason) +
  plot_layout(heights = c(1, 1.15)) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(face = "bold", size = 16))

plotTipoMuestreoPunto2

ggsave("Figure1.jpg", plot = plotTipoMuestreoPunto2, width = 8, height = 7.5, dpi = 300)


ggsave("Figure1B.jpg", plot = plotTipoMuestreoPunto2, width = 9, height = 7.5, dpi = 300)


ggsave("Figure1.pdf", plot = plotTipoMuestreoPunto2, width = 8, height = 7.5, dpi = 300)
```

Statistical tests
```{r, eval=FALSE}
library(tidyverse)
library(tibble)
library(FSA)

# --- SEDIMENT, DRY SEASON ---

# Estimate alpha diversity metrics for sediment samples from the dry season
alphadiv <- estimate_richness(Sediment2_Seca2, measures = c("Observed", "Shannon", "InvSimpson"))
alphadiv2 <- rownames_to_column(alphadiv, var = "Codigo") 
meta_data2 <- meta_data
alphadiv3 <- left_join(alphadiv2, meta_data2, by = "Codigo")

# Perform Kruskal-Wallis test and Dunn post-hoc test for each diversity metric by sampling site (Punto)
kruskal.test(Observed ~ Punto, data = alphadiv3) # p-value = 0.3012
dunnTest(Observed ~ Punto, data = alphadiv3, method = "bonferroni") #no significant

kruskal.test(Shannon ~ Punto, data = alphadiv3) #p-value = 0.02732
dunnTest(Shannon ~ Punto, data = alphadiv3, method = "bonferroni")
#1 Site  1 - Site  2 -1.341641 0.179712495 0.53913748
#2 Site  1 - Site  3 -2.683282 0.007290358 0.02187107 significant
#3 Site  2 - Site  3 -1.341641 0.179712495 0.53913748

kruskal.test(InvSimpson ~ Punto, data = alphadiv3)#p-value = 0.03899
dunnTest(InvSimpson ~ Punto, data = alphadiv3, method = "bonferroni")
#1 Site  1 - Site  2 -1.490712 0.1360371 0.40811138
#2 Site  1 - Site  3 -2.534210 0.0112701 0.03381031 significant
#3 Site  2 - Site  3 -1.043498 0.2967175 0.89015258


# --- WATER COLUMN, DRY SEASON ---

# Estimate alpha diversity metrics for water samples from the dry season
alphadiv5 <- estimate_richness(ColaAgua_Seca2, measures = c("Observed", "Shannon", "InvSimpson"))
alphadiv6 <- rownames_to_column(alphadiv5, var = "Codigo") 
alphadiv7 <- left_join(alphadiv6, meta_data, by = "Codigo")

# Run Kruskal-Wallis and Dunn's test by site (Punto)
kruskal.test(Observed ~ Punto, data = alphadiv7) #p-value = 0.02732
dunnTest(Observed ~ Punto, data = alphadiv7, method = "bonferroni")
#1 Site  1 - Site  2 -1.341641 0.179712495 0.53913748
#2 Site  1 - Site  3 -2.683282 0.007290358 0.02187107 significant
#3 Site  2 - Site  3 -1.341641 0.179712495 0.53913748

kruskal.test(Shannon ~ Punto, data = alphadiv7) #p-value = 0.02732
dunnTest(Shannon ~ Punto, data = alphadiv7, method = "bonferroni")
#1 Site  1 - Site  2 -1.341641 0.179712495 0.53913748
#2 Site  1 - Site  3 -2.683282 0.007290358 0.02187107 significant
#3 Site  2 - Site  3 -1.341641 0.179712495 0.53913748

kruskal.test(InvSimpson ~ Punto, data = alphadiv7)#p-value = 0.06081
dunnTest(InvSimpson ~ Punto, data = alphadiv7, method = "bonferroni")#non significant


# --- WATER COLUMN, RAINY SEASON ---

# Estimate diversity metrics for water samples from the rainy season
alphadiv8 <- estimate_richness(ColaAgua_Lluvioso2, measures = c("Observed", "Shannon", "InvSimpson"))
alphadiv9 <- rownames_to_column(alphadiv8, var = "Codigo") 
alphadiv10 <- left_join(alphadiv9, meta_data, by = "Codigo")

# Run statistical tests
kruskal.test(Observed ~ Punto, data = alphadiv10) #p-value = 0.06646
dunnTest(Observed ~ Punto, data = alphadiv10, method = "bonferroni")#non significant

kruskal.test(Shannon ~ Punto, data = alphadiv10) #p-value = 0.02732
dunnTest(Shannon ~ Punto, data = alphadiv10, method = "bonferroni")
#1 Site  1 - Site  2 1.341641 0.179712495 0.53913748
#2 Site  1 - Site  3 2.683282 0.007290358 0.02187107 significant
#3 Site  2 - Site  3 1.341641 0.179712495 0.53913748

kruskal.test(InvSimpson ~ Punto, data = alphadiv10) #p-value = 0.06646
dunnTest(InvSimpson ~ Punto, data = alphadiv10, method = "bonferroni") #non significant


# --- SEDIMENT, RAINY SEASON ---

# Estimate diversity metrics for sediment samples from the rainy season
alphadiv11 <- estimate_richness(Sediment2_Lluvioso2, measures = c("Observed", "Shannon", "InvSimpson"))
alphadiv12 <- rownames_to_column(alphadiv11, var = "Codigo") 
alphadiv13 <- left_join(alphadiv12, meta_data, by = "Codigo")

# Run Kruskal-Wallis and post-hoc Dunn tests for diversity metrics
kruskal.test(Observed ~ Punto, data = alphadiv13)#p-value = 0.8712
dunnTest(Observed ~ Punto, data = alphadiv13, method = "bonferroni")#non significant

kruskal.test(Shannon ~ Punto, data = alphadiv13)#p-value = 0.3932
dunnTest(Shannon ~ Punto, data = alphadiv13, method = "bonferroni")#non significant

kruskal.test(InvSimpson ~ Punto, data = alphadiv13)#p-value = 0.2019
dunnTest(InvSimpson ~ Punto, data = alphadiv13, method = "bonferroni")#non significant
```

Beta Diversity analyses
```{r}
# --- AGGREGATE AND FILTER TAXA BY CLASS LEVEL ---
library(ggplot2)
library(phyloseq)
library(ggh4x)


# Okabe-Ito hex codes for 6 classes:
okabe_ito_colors <- c(
  "#CC79A7",  # Redish purple
  "#E69F00", # Orange
  "#56B4E9", # Sky Blue
  "#009E73", # Bluish Green
  "#F0E442", # Yellow
  "#0072B2" # Blue
)

# Apply to your scale_fill_manual:
scale_fill_manual(values = okabe_ito_colors)


# 1. Melt phyloseq object to a standard data frame
df_plot <- psmelt(ps_rel_abundclass)

# 2. Correctly assign 'Water' vs 'Sediment'
df_plot$Tipo_Name <- ifelse(df_plot$Tipo == "A" | grepl("^A", df_plot$Sample), "Water", "Sediment")

# 3. Format season labels
df_plot$Muestreo_Label <- factor(df_plot$Muestreo, 
                                 levels = c("2", "3"), 
                                 labels = c("Dry season", "Rainy season"))

# 4. REPLACE "Sitio" WITH "Site" IN PUNTO
df_plot$Punto_Label <- gsub("Sitio", "Site", df_plot$Punto)

class_Estacion2 <- ggplot(df_plot, aes(x = Sample, y = Abundance, fill = class)) +
  # Thin white borders between segments; removes colored outlines entirely
  geom_bar(stat = "identity", position = "stack", width = 0.85, color = "white", linewidth = 0.2) +
  
  # NESTED FACETING: Season on top, Sitio in middle, Tipo (Water/Sediment) at bottom
  facet_nested(
    ~ Muestreo_Label + Punto_Label + Tipo_Name, 
    scales = "free_x", 
    space = "free_x",
    nest_line = element_line(color = "white", linewidth = 0.8), # Clean underline annotation
    solo_line = TRUE
  ) + 
  
  # Colors
  #scale_fill_manual(values = setNames(combined_colors[1:length(class_names)], class_names)) +
  scale_fill_manual(values = okabe_ito_colors) +
  scale_y_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
  
  labs(
    y = "Relative abundance",  
    fill = "Class"
  ) +
  
  theme_classic() + # Cleaner base theme without box borders
  theme(
    # Background & Panels
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_blank(),
    
    # X & Y Axis Text
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = "black", face = "bold", size = 9), 
    axis.text.y = element_text(color = "black", size = 10), 
    #axis.ticks.x = element_blank(), # Removes tick marks for cleaner spacing
    #axis.ticks.y = element_line(color = "black"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 13, face = "bold"),
    
    # Header Boxes Styling (Matches reference image style)
    strip.background = element_rect(fill = "grey85", color = NA), # No outline around strip boxes
    strip.text = element_text(size = 9, face = "bold", color = "black", margin = margin(t = 4, b = 4)),
    panel.spacing.x = unit(0.3, "lines"), # White gap between facet panels
    
    # Underline styling for ggh4x
    ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.6),
    
    # Legend
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10)
  )

# Render plot
class_Estacion2



# --- AGGREGATE AND FILTER TAXA BY ORDER LEVEL ---

# Group taxa at the "order" level (aggregates ASVs/OTUs into taxonomic orders)
physeqOrder = tax_glom(p.rar4.2, taxrank = "order")
# Transform count data into relative abundances per sample
physeqOrderRA = transform_sample_counts(physeqOrder, function(x) x / sum(x))
# Filter out orders with mean relative abundance ≤ 1%
physeqOrderRAF = filter_taxa(physeqOrderRA, function(x) mean(x) > 0.01, TRUE)
# Get the names of taxonomic orders that passed the abundance filter
keepOrders = get_taxa_unique(physeqOrderRAF, "order")
# Subset the original dataset to retain only taxa belonging to the filtered orders
physeqF = subset_taxa(p.rar4.2, order %in% keepOrders)
physeqF = tax_glom(physeqF, taxrank = "order")
# Convert counts in the filtered dataset to relative abundances
ps_rel_abund = transform_sample_counts(physeqF, function(x) x / sum(x))

# Create a color palette (choose one you like)
library(RColorBrewer)
custom_colors <- c(
  brewer.pal(8, "Set2"),
  brewer.pal(8, "Dark2"),
  "#999999", "#E0E0E0"  # Add more if needed
)

# Get unique order names (sorted for consistency)
order_names <- sort(unique(as.character(tax_table(ps_rel_abund)[, "order"])))


ps_rel_abund@tax_table[, "order"] <- as.character(ps_rel_abund@tax_table[, "order"])

# Create a stacked bar plot showing the relative abundance of taxonomic orders per sample
orden_Estacion <- plot_bar(ps_rel_abund, fill = "order") +
  geom_bar(aes(color = as.factor(order), fill = as.factor(order)), stat = "identity", position = "stack") +
  labs(
    x = "Samples",
    y = "Relative abundance ",
    fill = "Order",
    color = "Order"
  ) +
  facet_wrap(~ Muestreo, scales = "free",
             labeller = labeller(Muestreo = c(
               "2" = "Dry season",
               "3" = "Raining season"
             ))) +
  
  scale_fill_manual(values = setNames(custom_colors[1:length(order_names)], order_names)) +
  scale_color_manual(values = setNames(custom_colors[1:length(order_names)], order_names)) +
  theme(
    panel.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold", colour = "black"),
    axis.text.y = element_text(color = "black"), 
    axis.ticks.x = element_line(),
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold")
  ) +
  guides(
    fill = guide_legend(title = "Order"),
    color = guide_legend(title = "Order")
  )

orden_Estacion


#AQUIIII ES
library(ggplot2)
library(phyloseq)
library(ggh4x)

# Okabe-Ito hex codes for 6 classes:
okabe_ito_colors <- c(
  "#CC79A7",  # Redish purple
  "#E69F00", # Orange
  "#56B4E9", # Sky Blue
  "#009E73", # Bluish Green
  "#F0E442", # Yellow
  "#0072B2" # Blue
)


# 10 publication-ready colorblind-safe hex codes
cvd_10_colors <- c(
  "#56B4E9", # Cyan / Light Blue
  "#D55E00", # Rose / Muted Red
  "#F0E442", # Sand / Light Yellow
  "#009E73", # Dark Green
  "#332288", # Dark Indigo / Purple
  "#CC79A7", # Magenta
  "#0072B2", # Teal
  "#999933", # Olive
  "#882255", # Wine / Dark Rose
  "#E69F00"  # Orange
)

# 1. Melt phyloseq object to a standard data frame
df_order <- psmelt(ps_rel_abund)

# 2. Correctly assign 'Water' vs 'Sediment'
df_order$Tipo_Name <- ifelse(df_order$Tipo == "A" | grepl("^A", df_order$Sample), "Water", "Sediment")

# 3. Format season labels
df_order$Muestreo_Label <- factor(df_order$Muestreo, 
                                  levels = c("2", "3"), 
                                  labels = c("Dry season", "Rainy season"))

# 4. REPLACE "Sitio" WITH "Site" IN PUNTO
df_order$Punto_Label <- gsub("Sitio", "Site", df_order$Punto)

# 4. Build plot matching class_Estacion2 format
orden_Estacion2 <- ggplot(df_order, aes(x = Sample, y = Abundance, fill = order)) +
  # White borders between segments without outer colored outlines
  geom_bar(stat = "identity", position = "stack", width = 0.85, color = "white", linewidth = 0.2) +
  
  # NESTED FACETING: Season on top, Sitio in middle, Tipo (Water/Sediment) at bottom
  facet_nested(
    ~ Muestreo_Label + Punto_Label + Tipo_Name, 
    scales = "free_x", 
    space = "free_x",
    nest_line = element_line(color = "white", linewidth = 0.8), # Clean underline annotation
    solo_line = TRUE
  ) + 
  
  # Colors & Scales
  #scale_fill_manual(values = setNames(custom_colors[1:length(order_names)], order_names)) +
  scale_fill_manual(values = cvd_10_colors[1:length(order_names)]) +
  scale_y_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
  
  labs(
    y = "Relative abundance",  
    fill = "Order"
  ) +
  
  theme_classic() + 
  theme(
    # Background & Panel
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    
    # REMOVE AXIS LINES
    axis.line = element_blank(),
    
    # Text & Ticks
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = "black", face = "bold", size = 9), 
    axis.text.y = element_text(color = "black", size = 10), 
    axis.ticks.x = element_blank(),     # Removes x ticks for clean layout
    axis.ticks.y = element_line(color = "black"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 13, face = "bold"),
    
    # Header Boxes Styling
    strip.background = element_rect(fill = "grey85", color = NA), 
    strip.text = element_text(size = 9, face = "bold", color = "black", margin = margin(t = 4, b = 4)),
    panel.spacing.x = unit(0.3, "lines"), 
    
    # Underline styling for ggh4x
    ggh4x.facet.nestline = element_line(color = "black", linewidth = 0.6),
    
    # Legend
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10)
  )

# Render plot
orden_Estacion2


# Save the Plot
ggsave("RelativeAbundanceOrder.jpg", plot = orden_Estacion, width = 10, height = 7, dpi = 300)
ggsave("RelativeAbundanceOrder.pdf", plot = orden_Estacion, width = 10, height = 7, dpi = 300)

# Arrange the class-level and order-level plots in one figure (stacked vertically)
plotrelativeabundance = grid.arrange(class_Estacion2, orden_Estacion2, ncol = 1)

plotrelativeabundance <- (class_Estacion2 / orden_Estacion2) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(face = "bold", size = 16))

plotrelativeabundance

# Save the combined plot as a jpg file
install.packages("gridExtra")
library(gridExtra)

ggsave("plotrelativeabundance.jpg", plot = plotrelativeabundance, width = 15, height = 12, dpi = 450)

```

ORDINATION PLOT
```{r}
library(ggplot2)
library(phyloseq)
library(microbiome)
library(ggforce)
library(ggrepel)

# 1. Prepare metadata
sample_data(p.rar4.2)$Muestreo <- as.factor(sample_data(p.rar4.2)$Muestreo)

df_ord <- plot_ordination(p.rar4.2, ord_clr, type = "samples", justDF = TRUE)
df_ord$Site_Num <- gsub(".*([0-9]+).*", "\\1", df_ord$Punto)

# Format factor for Site legend labels
df_ord$Site <- factor(df_ord$Site_Num, levels = c("1", "2", "3"), labels = c("Site 1", "Site 2", "Site 3"))

# ES AQUIIIIIIIII

library(ggplot2)
library(phyloseq)
library(microbiome)
library(ggforce)
library(ggrepel)

# 1. Prepare metadata
sample_data(p.rar4.2)$Muestreo <- as.factor(sample_data(p.rar4.2)$Muestreo)

df_ord <- plot_ordination(p.rar4.2, ord_clr, type = "samples", justDF = TRUE)
df_ord$Site_Num <- gsub(".*([0-9]+).*", "\\1", df_ord$Punto)

# Format factor for Site legend labels
df_ord$Site <- factor(df_ord$Site_Num, levels = c("1", "2", "3"), labels = c("Site 1", "Site 2", "Site 3"))

# 2. Build Plot
ordination_plotp <- ggplot(df_ord, aes(x = PC1, y = PC2)) +
  
  # Ellipses
  ggforce::geom_mark_ellipse(
    aes(group = Muestreo, color = Muestreo), 
    alpha = 0.3, 
    linewidth = 0.8, 
    linetype = "dashed"
  ) +
  
  # Grid lines
  geom_hline(yintercept = c(-7, 0, 7), color = "grey85", linetype = "dashed") +
  geom_vline(xintercept = c(-7, 0, 7), color = "grey85", linetype = "dashed") +
  
  # POINT SHAPES
  geom_point(
    aes(color = Muestreo, shape = Tipo), 
    size = 3.5, 
    alpha = 0.85
  ) +
  
  # NUMBERS ON PLOT + SITE LEGEND MAPPING
  geom_text_repel(
    aes(label = Site_Num, alpha = Site),
    size = 3,
    fontface = "bold",
    color = "black",
    box.padding = 0.25,
    point.padding = 0.2,
    max.overlaps = Inf,
    segment.color = NA
  ) +
  
  # Aspect ratio & limits
  # coord_fixed(ratio = clr2 / clr1, xlim = c(-7, 7), ylim = c(-7, 7)) +
  coord_cartesian(xlim = c(-7, 7), ylim = c(-7, 7)) +
  
  # Color Scale (Sampling) -> Changed to "Dry" and "Rainy"
  scale_color_manual(
    values = c("2" = "#CC79A7", "3" = "#009E73"),
    labels = c("2" = "Dry", "3" = "Rainy"),
    name = "Sampling"
  ) +
  
  # Shape Scale (Matrix) -> Changed title to "Matrix"
  scale_shape_manual(
    values = c("sedimento" = 16, "agua" = 17),
    labels = c("sedimento" = "Sediment", "agua" = "Water"),
    name = "Matrix"
  ) +
  
  # Alpha Scale (Site legend)
  scale_alpha_manual(
    values = c("Site 1" = 1, "Site 2" = 1, "Site 3" = 1),
    name = "Site"
  ) +
  
  # Customize legend glyphs for Site
  guides(
    alpha = guide_legend(
      override.aes = list(
        label = c("1", "2", "3"),
        size = 4,
        color = "black"
      )
    )
  ) +
  
  labs(
    x = paste0("PC1 (", round(clr1 * 100, 0), "%)"),
    y = paste0("PC2 (", round(clr2 * 100, 0), "%)")
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_blank(),  
    axis.line = element_blank(),     
    axis.ticks = element_blank(),
    
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    
    # Bottom-right Legend box
    legend.position = c(0.95, 0.05),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.box = "vertical",
    legend.text = element_text(size = 9), 
    legend.title = element_text(size = 10, face = "bold"), 
    legend.key.size = unit(0.35, "cm") 
  )

# Render plot
ordination_plotp


# 2. Build Plot
ordination_plotp <- ggplot(df_ord, aes(x = PC1, y = PC2)) +
  
  # Ellipses
  ggforce::geom_mark_ellipse(
    aes(group = Muestreo, color = Muestreo), 
    alpha = 0.3, 
    linewidth = 0.8, 
    linetype = "dashed"
  ) +
  
  # Grid lines aligned with narrowed bounds
  geom_hline(yintercept = c(-4, 0, 4), color = "grey85", linetype = "dashed") +
  geom_vline(xintercept = c(-4, 0, 4), color = "grey85", linetype = "dashed") +
  
  # POINT SHAPES
  geom_point(
    aes(color = Muestreo, shape = Tipo), 
    size = 3.5, 
    alpha = 0.85
  ) +
  
  # NUMBERS ON PLOT + SITE LEGEND MAPPING
  geom_text_repel(
    aes(label = Site_Num, alpha = Site),
    size = 3.5,            # Increased text size slightly for clarity
    fontface = "bold",
    color = "black",
    box.padding = 0.25,
    point.padding = 0.25,
    max.overlaps = Inf,
    segment.color = NA
  ) +
  
  # Tightened limits to reduce white space + fixed aspect ratio
  coord_fixed(
    ratio = 1,             # Keeps equal scaling for PC units
    xlim = c(-5, 5),       # Zoomed in from -7, 7
    ylim = c(-5, 5)        # Zoomed in from -7, 7
  ) +
  
  # Color Scale (Sampling)
  scale_color_manual(
    values = c("2" = "#CC79A7", "3" = "#009E73"),
    labels = c("2" = "Dry", "3" = "Rainy"),
    name = "Sampling"
  ) +
  
  # Shape Scale (Matrix)
  scale_shape_manual(
    values = c("sedimento" = 16, "agua" = 17),
    labels = c("sedimento" = "Sediment", "agua" = "Water"),
    name = "Matrix"
  ) +
  
  # Alpha Scale (Site legend)
  scale_alpha_manual(
    values = c("Site 1" = 1, "Site 2" = 1, "Site 3" = 1),
    name = "Site"
  ) +
  
  # Customize legend glyphs for Site
  guides(
    alpha = guide_legend(
      override.aes = list(
        label = c("1", "2", "3"),
        size = 4,
        color = "black"
      )
    )
  ) +
  
  labs(
    x = paste0("PC1 (", round(clr1 * 100, 0), "%)"),
    y = paste0("PC2 (", round(clr2 * 100, 0), "%)")
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_blank(),  
    axis.line = element_blank(),     
    axis.ticks = element_blank(),
    
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    
    # Bottom-right Legend box
    legend.position = c(0.95, 0.05),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    legend.box = "vertical",
    legend.text = element_text(size = 9), 
    legend.title = element_text(size = 10, face = "bold"), 
    legend.key.size = unit(0.35, "cm") 
  )

# Render plot
ordination_plotp

# Save the ordination plot as a jpg file
ggsave("ordinationplot.jpg", plot = ordination_plotp, width = 15, height = 7, dpi = 300)


ggsave("Figure3.jpg", plot = ordination_plotp, width = 8, height = 7, dpi = 300)


# Save the ordination plot as a pdf file with wider width
ggsave("Figure3.pdf", plot = ordination_plotp, width = 16, height = 7, dpi = 300)


#PERMANOVA
install.packages("vegan")
library(vegan)

#Extract the OTU table from p_clr
otu_table(p_clr)
otu_clr <- as(otu_table(p_clr), "matrix")

#Transposase row x columns
if(taxa_are_rows(p_clr)){
  otu_clr <- t(otu_clr)
}

#Compute Euclidean Distance (correct for CLR)
dist_clr <- dist(otu_clr, method = "euclidean")

#Extract Metadata
meta <- as(sample_data(p_clr), "data.frame")

#Run PERMANOVA
adonis2(dist_clr ~ Muestreo + Tipo + Punto,
        data = meta,
        permutations = 999)



adonis2(dist_clr ~ Muestreo + Tipo + Punto,
        data = meta,
        permutations = 999,
        by = "margin")
```

COG analysis

SAMPLING
```{r}
summary(Sampling)

Sampling$Sampling=as.factor(Sampling$Sampling)
Sampling$COG=as.factor(Sampling$COG)
Sampling$Function=as.factor(Sampling$Function)
summary(Sampling)



#Filter dataset
Sampling_clean <- Sampling %>% 
  filter(COG != "Not_assigned")


# Recode Sampling values
Combined_pct <- Combined_pct %>%
  mutate(Season = recode(Sampling,
                         `2` = "Dry season",
                         `3` = "Rainy season"))

Combined_pct2 <- Combined_pct2 %>% 
  filter(COG != "S")

Combined_pct2 <- Combined_pct2 %>% 
  filter(Count != "0")

Combined_pct2 <- Combined_pct2 %>%
  filter(as.numeric(Count) > 1)

Combined_pct2 <- Combined_pct2 %>% 
  filter(COG != "Z")

Combined_pct2 <- Combined_pct2 %>% 
  filter(COG != "A")

Combined_pct2 <- Combined_pct2 %>% 
  filter(COG != "B")


# 1. Clean unassigned reads and calculate percentages on the full dataset
Combined_pct <- Sampling %>% 
  filter(COG != "Not_assigned") %>%
  group_by(Sampling) %>%
  mutate(Percent = (Count / sum(Count)) * 100) %>%
  ungroup() %>%
  mutate(Season = recode(Sampling,
                         `2` = "Dry season",
                         `3` = "Rainy season"))

# 2. Filter out specific categories and low counts from the percentage dataset
Combined_pct2 <- Combined_pct %>% 
  filter(!COG %in% c("S", "Z", "A", "B")) %>%
  filter(Count != "0", as.numeric(Count) > 1) %>%
  mutate(COG = fct_reorder(COG, Percent, .fun = mean, .desc = TRUE))

# 3. Generate Plot
bar_plot_Sampling2 <- ggplot(Combined_pct2, aes(x = COG, y = Percent, fill = Season)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9),
    width = 0.7,
    color = "black"
  ) +
  scale_fill_manual(
    values = c("Dry season" = "#CC79A7", "Rainy season" = "#009E73")  

  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25)) +
  labs(
    x = "COG Categories",
    y = "Percentage of COG (%)",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top",
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    axis.text = element_text(size = 11)
  )

bar_plot_Sampling2

#Percentages S: Dry:35 Raining:43
#Removed COG equal or less to 0.1% jeje

#  "A: RNA processing and modification",
 # "B: Chromatin structure and dynamics",
#  "R: General function prediction only",
#  "W: Extracellular structures",
#  "Y: Nuclear structure",
 # "Z: Cytoskeleton",

cog_functions <- paste(
  "C: Energy production and conversion",
  "D: Cell cycle control, cell division, chromosome partitioning",
  "E: Amino acid transport and metabolism",
  "F: Nucleotide transport and metabolism",
  "G: Carbohydrate transport and metabolism",
  "H: Coenzyme transport and metabolism",
  "I: Lipid transport and metabolism",
  "J: Translation, ribosomal structure and biogenesis",
  "K: Transcription",
  "L: Replication, recombination and repair",
  "M: Cell wall/membrane/envelope biogenesis",
  "N: Cell motility",
  "O: Posttranslational modification, protein turnover, chaperones",
  "P: Inorganic ion transport and metabolism",
  "Q: Secondary metabolites biosynthesis, transport and catabolism",
  "T: Signal transduction mechanisms",
  "U: Intracellular trafficking, secretion, and vesicular transport",
  "V: Defense mechanisms",
  sep = "\n"
)

final_plot_Sampling <- ggdraw() +
  draw_plot(bar_plot_Sampling2, x = 0, y = 0, width = 0.75, height = 1) +
  draw_text(cog_functions, x = 0.63, y = 0.89, hjust = 0, vjust = 1, size = 11, lineheight = 1.5)

final_plot_Sampling


ggsave("Figure4.jpg", final_plot_Sampling, width = 12, height = 7, dpi = 300)

ggsave("Figure4.pdf", final_plot_Sampling, width = 13, height = 7, dpi = 450)


# Show the final plot
print(final_plot_Sampling)


```

TYPE
```{r}

summary(Type)

Type$Type=as.factor(Type$Type)
Type$COG=as.factor(Type$COG)
Type$Function=as.factor(Type$Function)
summary(Type)

#Filter dataset
Type_clean <- Type %>% 
  filter(COG != "Not_assigned")

#AQUIIII
Combined_pct3 <- Type_clean %>%
  group_by(Type) %>%
  mutate(Percent = (Count / sum(Count)) * 100)


Combined_pct3 <- Combined_pct3 %>% 
  filter(COG != "S")

Combined_pct3 <- Combined_pct3 %>% 
  filter(Count != "0")

Combined_pct3 <- Combined_pct3 %>%
  filter(as.numeric(Count) > 1)

Combined_pct3 <- Combined_pct3 %>% 
  filter(COG != "Z")

Combined_pct3 <- Combined_pct3 %>% 
  filter(COG != "A")

Combined_pct3 <- Combined_pct3 %>% 
  filter(COG != "B")

library(forcats)

Combined_pct3 <- Combined_pct3 %>%
  mutate(COG = fct_reorder(COG, Percent, .fun = mean, .desc = TRUE))


summary(Combined_pct3)

 bar_plot_Type <- ggplot(Combined_pct3, aes(x = COG, y = Percent, fill = Type)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9),
    width = 0.7,
    color = "black",
  ) +
  scale_fill_manual(
    values = c("sediment" = "#E69F00", "water" = "#56B4E9")  
  ) +
  theme_minimal() +
  labs(
    x = "COG Categories",
    y = "Percentage of COG (%)",
    fill = NULL
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25)) +  
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top",
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    axis.text = element_text(size = 10)
  )
bar_plot_Type

final_plot_Type <- ggdraw() +
  draw_plot(bar_plot_Type, x = 0, y = 0, width = 0.75, height = 1) +
  draw_text(cog_functions, x = 0.67, y = 0.89, hjust = 0, vjust = 1, size = 9.5, lineheight = 1.5)

final_plot_Type

ggsave("COG_Type.jpg", final_plot_Type, width = 13, height = 7, dpi = 450)
ggsave("COG_Type.pdf", final_plot_Type, width = 13, height = 7, dpi = 450)


# Show the final plot
print(final_plot_Type)
```

POINT
```{r}
summary(Point)

Point$Point=as.factor(Point$Point)
Point$COG=as.factor(Point$COG)
Point$Function=as.factor(Point$Function)
summary(Point)

Point_clean <- Point %>% 
  filter(COG != "Not_assigned")

#AQUIIII
Combined_pct4 <- Point_clean %>%
  group_by(Point) %>%
  mutate(Percent = (Count / sum(Count)) * 100)


Combined_pct4 <- Combined_pct4 %>% 
  filter(COG != "S")

Combined_pct4 <- Combined_pct4 %>% 
  filter(Count != "0")

Combined_pct4 <- Combined_pct4 %>%
  filter(as.numeric(Count) > 1)

Combined_pct4 <- Combined_pct4 %>% 
  filter(COG != "Z")

Combined_pct4 <- Combined_pct4 %>% 
  filter(COG != "A")

Combined_pct4 <- Combined_pct4 %>% 
  filter(COG != "B")

Combined_pct4 <- Combined_pct4 %>% 
  filter(COG != "W")

library(forcats)

Combined_pct4 <- Combined_pct4 %>%
  mutate(COG = fct_reorder(COG, Percent, .fun = mean, .desc = TRUE))


summary(Combined_pct4)

bar_plot_Point <- ggplot(Combined_pct4, aes(x = COG, y = Percent, fill = Point)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 1),
    width = 0.5,
    color = "black"
  ) +
  scale_fill_manual(
    values = c("Site 1" = "#CC79A7", "Site 2" = "#009E73", "Site 3" = "#56B4E9")   
  ) +
  theme_minimal() +
  labs(
    x = "COG Categories",
    y = "Percentage of COG (%)",
    fill = NULL
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25)) +  
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top",
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    axis.text = element_text(size = 10)
  )
bar_plot_Point

final_plot_Point <- ggdraw() +
  draw_plot(bar_plot_Point, x = 0, y = 0, width = 0.75, height = 1) +
  draw_text(cog_functions, x = 0.67, y = 0.89, hjust = 0, vjust = 1, size = 9.5, lineheight = 1.5)

final_plot_Point

ggsave("COG_PointFINAL.jpg", final_plot_Point, width = 13, height = 7, dpi = 450)
ggsave("COG_PointFINAL_SU.pdf", final_plot_Point, width = 13, height = 7, dpi = 450)


install.packages("patchwork")
library(patchwork)


# Define a fixed factor order based on overall mean abundance
cog_order <- c("L", "M", "K", "T", "F", "O", "H", "E", "J", "G", "V", "C", "D", "P", "I", "U", "Q", "N")

Combined_pct3$COG <- factor(Combined_pct3$COG, levels = cog_order)
Combined_pct4$COG <- factor(Combined_pct4$COG, levels = cog_order)

# 3. Stack vertically and add panel tags (A and B)
Supplementary_Fig1 <- (bar_plot_Type / bar_plot_Point) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(face = "bold", size = 16))


final_Supplementary_Fig1 <- ggdraw() +
  draw_plot(Supplementary_Fig1, x = 0, y = 0, width = 0.72, height = 1) +
  draw_text(
    cog_functions, 
    x = 0.72, 
    y = 0.82, 
    hjust = 0, 
    vjust = 1, 
    size = 8.5, 
    lineheight = 1.25
  )

# Display final combined figure
final_Supplementary_Fig1

ggsave("SuplementaryFig1.jpg", final_Supplementary_Fig1, width = 13, height = 7, dpi = 450)
ggsave("SuplementaryFig1.pdf", final_Supplementary_Fig1, width = 13, height = 7, dpi = 450)


```


